**The difference between precision and accuracy in assigning 16S rRNA gene sequences to operational taxonomic units**

### 2015-09-29 Schloss_Cluster_PeerJ_2015.Rmd
Build figures 1 and 2

The growth in options for assigning sequences using each of these three broad approaches has been considerable. It has been difficult to objectively assess the quality of OTU assignments. Some have focused on the time and memory required to process a dataset. These are valid parameters to assess when judging a clustering algorithm, but have little to say about the quality of the clustering. Others have attempted to judge the quality of an algorithm by its ability to generate data that parallels classification data. This approach is problematic because bacterial taxonomy often reflects the biases amongst bacterial systematicists and the rates of evolution across lineages are not the same. We recently proposed an approach for evaluating OTU assignments using the similarity between sequences (REF). Those sequences that were similar to each other and found in the same OTU were called true positives while those that were similar and found in different OTUs were called false negatives. Meanwhile, those sequences that were different from each other and found in the same OTU were called false positives and those that were dissimilar and found in different OTUs were called true negatives. Counting the frequency of these different classes allowed us to judge how each method balanced the ratio of true positives and negatives to false positives and ratios using the Matthew's correlation coefficient (MCC; REF). That analysis focused on assessing *de novo* clustering algorithms and found that the average neighbor algorithm outperformed the other hierarchical and heuristic algorithms.

A recent analysis by He and colleagues (REF) attempted to characterize the three general clustering approaches by focusing on what they called stability. They defined stability as the ability of an algorithm to provide the same clustering on a subset of the data as was found in the full dataset. Related to this, the authors expressed concern that because some algorithms use a random number generator to break ties there may be further instability between executions of algorithm. Their concept of stability fails to quantify for the accuracy of the clustering and instead focuses on the precision of the clustering. A method may be very precise, but low in accuracy. In the current analysis, we attempted to assess the accuracy and precision of the various clustering algorithms. Building on our previous analysis of clustering algorithms, our hypothesis was that the algorithms praised by the He study for their stability actually suffered a lack of accuracy. In addition, we assess these parameters in light of sequence quality using the original 454 dataset and a larger and more modern dataset generated using the MiSeq platform.

@@

***Summary of He study.*** We sought to reproduce the more critical analyses performed in the He study. Similar to their study, we obtained the Canadian soil dataset from Roesch et al. (REF) and processed the sequences as described in their analysis. The first result we sought to reproduce was whether the clustering of the Roesch dataset using a subset of the data corresponded to the clustering observed with the full dataset. The He study found that when they used the open and closed-reference methods the subsampled clusters most closely resembled those of the full dataset when using the MCC. Among the *de novo* clustering methods they observed that the abundance-based greedy clustering (AGC) algorithm followed by single linkage (SL), distance-based greedy clustering (DGC), complete linkage (CL), and average linkage (AL) algorithms. They observed a broad range of MCC values among their AL replicates, which was not explained by the authors. The second result we sought to reproduce was the observation, originally made in the Roesch study, that rarefaction curves calculated using clusters obtained using a portion of the dataset did not overlap with rarefaction curves generated using clusters generated from the full dataset. In the He study, this appeared to be a more pronounced problem when using the CL, SL, and DGC algorithms.

***Clustering approaches vary in their precision.*** We first calculated the MCC for for each of the clustering algorithms using 20, 40, 60, and 80% relative to the clusters formed by the algorithms using the full dataset. Because a random number generator is used in some of algorithms to break ties where pairs of sequences have the same distance between them, similar to the He study, we replicated each algorithm and subsample 30 times.

***Effect of sampling depth on observed number of OTUs.***

@@

***Accuracy and precision affected by data quality.***

>>>>

The growth in options for assigning sequences using each of these three broad approaches has been considerable. It has been difficult to objectively assess the quality of OTU assignments. Some have focused on the time and memory required to process a dataset (REFS). These are valid parameters to assess when judging a clustering algorithm, but have little to say about the quality of the clustering. Others have attempted to judge the quality of an algorithm by its ability to generate data that parallels classification data. This approach is problematic because bacterial taxonomy often reflects the biases amongst bacterial systematicists and the rates of evolution across lineages are not the same. We recently proposed an approach for evaluating OTU assignments using the similarity between sequences (REF). Those sequences that were similar to each other and found in the same OTU were called true positives while those that were similar and found in different OTUs were called false negatives. Meanwhile, those sequences that were different from each other and found in the same OTU were called false positives and those that were dissimilar and found in different OTUs were called true negatives. Counting the frequency of these different classes allowed us to judge how each method balanced the ratio of true positives and negatives to false positives and ratios using the Matthew's correlation coefficient (MCC; REF). That analysis focused on assessing *de novo* clustering algorithms and found that the average neighbor algorithm outperformed the other hierarchical and heuristic algorithms.

A recent analysis by He and colleagues (REF) attempted to characterize the three general clustering approaches by focusing on what they called stability. They defined stability as the ability of an algorithm to provide the same clustering on a subset of the data as was found in the full dataset. Related to this, the authors expressed concern that because some algorithms use a random number generator to break ties there may be further instability between executions of algorithm. Their concept of stability does not account for the accuracy of the clustering and instead focuses on the precision of the clustering. A method may be very precise, but low in accuracy. In the current analysis, we attempted to assess the accuracy and precision of the various clustering algorithms. Building on our previous analysis of clustering algorithms, our hypothesis was that the algorithms praised by the He study for their stability actually suffered a lack of accuracy. In addition, we assess these parameters in light of sequence quality using the original 454 dataset and a larger and more modern dataset generated using the MiSeq platform.

@@

***Summary of He study.*** We sought to identify the more critical analyses performed in the He study. Similar to their study, we obtained the Canadian soil dataset from Roesch et al. (REF) and processed the sequences as described in their analysis. In our opinion there were three important tests. First, was to quantify whether the clustering observed for a subset of the data represented the same clustering that was found with the full dataset. The He study found that when they used the open and closed-reference methods clusters formed using the subset data most closely resembled those of the full dataset. Among the *de novo* clustering methods they observed that the abundance-based greedy clustering (AGC) algorithm followed by single linkage (SL), distance-based greedy clustering (DGC), complete linkage (CL), and average linkage (AL) algorithms. They observed a broad range of MCC values among their AL replicates, which was not explained by the authors. Second, rarefaction curves calculated using clusters obtained using a portion of the dataset did not overlap with rarefaction curves generated using clusters generated from the full dataset. This result was originally observed in the Roesch study using the complete linkage algorithm and was reproduced in the He study where this result was more pronounced problem when using the CL, SL, and DGC algorithms relative to the other algorithms. Third, the authors attempted to describe the effects of the clustering instability on comparisons of communities. Such an analysis does not make sense for the *de novo* or open-reference approaches since the labels of the OTUs are not consistent between executions of the clustering algorithms. To overcome this, the authors selected representative sequences from OTUs and merged OTUs across different sized libraries to link OTUs. Regardless, the justification for this analysis is specious as the OTU assignments are based on the data available in the dataset when the sequences are clustered and comparing assignments in this manner are irreconcilable. We will attempt to replicate the first two observations.

***Clustering approaches vary in their stability.*** We first sought to assess calculated the MCC for for each of the clustering algorithms using 20, 40, 60, and 80% relative to the clusters formed by the algorithms using the full dataset (Figure 1A). Because a random number generator is used in some of algorithms to break ties where pairs of sequences have the same distance between them, similar to the He study, we replicated each algorithm and subsample 30 times. Across these sequencing depths, we observed that the stability of the SL and CL algorithms were highly sensitive to sampling effort relative to the AL, AGC, and DGC algorithms (Figure 1A). Our results (Figure 1B) largely confirmed those of Figure 4C of He study with one notable exception. In the He study, when comparing the OTUs formed using the AL algorithm with 60% of the data, they observed a mean MCC value of approximately 0.63 (95% confidence interval between approximately 0.15 and 0.75). In contrast, we observed a mean value of `r round(z[z$method=="an" & z$fraction==0.6, "mean"], 2)` (95% confidence interval between `r round(z[z$method=="an" & z$fraction==0.6, "lci"], 2)` and `r round(z[z$method=="an" & z$fraction==0.6, "uci"], 2)`). This result suggests that the AL algorithm was far more stabile than indicated in the He study.

***Effect of sampling depth on observed number of OTUs.*** Next, we sought to determine whether the number of OTUs observed when clustering a subset of the data was the same as that observed when rarefying the full dataset back to the number of sequences in the subset. The analyses from the He and Roesch studies both compared the number of observed OTUs from the subset to the average number of expected OTUs from the full dataset instead of to the 95% confidence interval of what would be expected from the full dataset.

@@

***Stability, accuracy, and precision affected by data quality.***
